---
title: "The Lattice Plotting System in R The Lattice Plotting System in R"
author: "Linda Angulo Lopez"
date: "17/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Lattice Plotting System in R The Lattice Plotting System in R
-xyplot: this is the main function for creating scatterplots
-bwplot: box-and-whiskers plots (“boxplots”)
-histogram: histograms
-stripplot: like a boxplot but with actual points
-dotplot: plot dots on "violin strings"
-splom: scatterplot matrix; like pairs in base plotting system
-levelplot, contourplot: for plotting "image" data


```{r}
#install.packages('lattice')
#xyplot(y ~ x | f * g, data)
library(lattice)
library(datasets)
## Simple scatterplot
xyplot(Ozone ~ Wind, data = airquality)
```


```{r}
library(datasets)
library(lattice)
## Convert 'Month' to a factor variable
airquality <- transform(airquality, Month = factor(Month))
xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5, 1))
```
Lattice graphics functions return an object of class trellis and autoprints the object to the graphic device. 

```{r}
p <- xyplot(Ozone ~ Wind, data = airquality) ## Nothing happens!
print(p) ## Plot appears
```
Lattice functions have a panel function which controls what happens inside each panel of the plot.

```{r}
set.seed(10)
x <- rnorm(100)
f <- rep(0:1, each = 50)
y <- x + f - f * x + rnorm(100, sd = 0.5)
f <- factor(f, labels = c("Group 1", "Group 2"))
xyplot(y ~ x | f, layout = c(2, 1)) ## Plot with 2 panels
```


```{r}
## Custom panel function
xyplot(y ~ x | f, panel = function(x, y, ...) {
 panel.xyplot(x, y, ...) ## First call the default panel function for 'xyplot'
 panel.abline(h = median(y), lty = 2) ## Add a horizontal line at the median
})

```


```{r}
## Custom panel function
xyplot(y ~ x | f, panel = function(x, y, ...) {
 panel.xyplot(x, y, ...) ## First call default panel function
 panel.lmline(x, y, col = 2) ## Overlay a simple linear regression line
})
```

```{r}
set.seed(723645)                                   # Create example data
data <- data.frame(x = rnorm(500),
                   y = rnorm(500),
                   group = c(rep("A", 250),
                             rep("B", 250)))
head(data)                                         # Head of example data

```

```{r}
plat1 <- xyplot(y ~ x, data[data$group == "A", ])  # Create plots
plat2 <- xyplot(y ~ x, data[data$group == "B", ])

```




## Including Plots

You can also embed plots, for example:

```{r}

#install.packages("gridExtra")                      # Install gridExtra package
library("gridExtra")                               # Load gridExtra package
grid.arrange(plat1, plat2, ncol = 2)               # Apply grid.arrange
```

```{r}
grid.arrange(plat1, plat2,plat1, plat2,plat1, plat2, plat1, plat2, ncol = 4, nrow=2)
```


[Read the docs](https://cran.r-project.org/web/packages/lattice/lattice.pdf)
