---
title: "Base Plotting in R"
author: "Linda Angulo Lopez"
date: "04/09/2020"
output: html_document
---

There are three key plotting systems in R, the Base, which is a type of artist's palette, model,  a Lattice System, which allows to specify an entire plot specified by one function and the conditioning ggplot2 with mixed elements of Base and Lattice, this is a demo of the base plot.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Base Plotting System


```{r }
library(datasets)
data(cars)
with(cars, plot(speed, dist))


```

## Examples for screen device display

- histograms

```{r}
library(datasets)
hist(airquality$Ozone) ## Draw a new plot

```

- scatter plots

```{r}
library(datasets)
with(airquality, plot(Wind, Ozone))

```

- box plots

```{r}
library(datasets)
airquality <- transform(airquality, Month = factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = "Month", ylab = "Ozone (ppb)")
```

- add the title

```{r}
library(datasets)
with(airquality, plot(Wind, Ozone))
title(main = "Ozone and Wind in New York City") ## Add a title
```

- use colour to symbolise

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
```


- Annotate

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City",
 type = "n"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
with(subset(airquality, Month != 5), points(Wind, Ozone, col = "red"))
legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other Months"))

```

-Regression Line

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City",
 pch = 20))
model <- lm(Ozone ~ Wind, airquality)
abline(model, lwd = 2)

```


- Multiple Plots

```{r}
par(mfrow = c(1, 2))
with(airquality, {
 plot(Wind, Ozone, main = "Ozone and Wind")
 plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
})

```


- more plots

```{r}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0))
with(airquality, {
 plot(Wind, Ozone, main = "Ozone and Wind")
 plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
 plot(Temp, Ozone, main = "Ozone and Temperature")
 mtext("Ozone and Weather in New York City", outer = TRUE)
})

```


#[by Linda, September 2020 ](https://www.linkedin.com/in/lindangulopez/)
