---
title: "Colours in R"
author: "Linda Angulo Lopez"
date: "20/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## colorRamp:

```{r}
 pal <- colorRamp(c("red", "blue"))
red <- pal(0)
blue <- pal(1)

red; blue

pal(seq(0, 1, len = 10))
```


## colorRampPalette:

```{r}
 pal <- colorRampPalette(c("red", "yellow"))
#red=FF0000; green=FFFF00; blue=FFFFFF

pal(10)
```


```{r}
library(RColorBrewer)
cols <- brewer.pal(100, "BuGn")
pal <- colorRampPalette(cols)
image(volcano, col = pal(15))
 
```


```{r}
image(volcano)

```

```{r}
x<- rnorm(1000)
y<-(2*x-x) + rnorm(1000)
smoothScatter(x,y)
```

```{r}

x<- rnorm(500)
y<-(2*x-x*x) + rnorm(150)
plot(x,y,col=rgb(0,0,0,.2), pch=19)
```



## library(grDevices)


```{r}
library(RColorBrewer)
cols <- brewer.pal(100, "BuGn")
pal <- colorRampPalette(cols)
image(volcano, col = pal(150))

```

```{r}
cols <- brewer.pal(100, "Spectral")
pal <- colorRampPalette(cols)
image(volcano, col = pal(150))

```
## Colors (ggplot2)

- use the colorblind-friendly palette 

```{r}

# Two variables
df <- read.table(header=TRUE, text='
 cond yval
    A 2
    B 2.5
    C 1.6
')

# Three variables
df2 <- read.table(header=TRUE, text='
 cond1 cond2 yval
    A      I 2
    A      J 2.5
    A      K 1.6
    B      I 2.2
    B      J 2.4
    B      K 1.2
    C      I 1.7
    C      J 2.3
    C      K 1.9
')
```

- Simple color assignment

```{r}

library(ggplot2)
# Default: dark bars
ggplot(df, aes(x=cond, y=yval)) + geom_bar(stat="identity")
# Bars with red outlines
ggplot(df, aes(x=cond, y=yval)) + geom_bar(stat="identity", colour="#FF9999") 
```



```{r}

# Red fill, black outlines
ggplot(df, aes(x=cond, y=yval)) + geom_bar(stat="identity", fill="#FF9999", colour="black")

```


```{r}
# Standard black lines and points
ggplot(df, aes(x=cond, y=yval)) + 
    geom_line(aes(group=1)) +     # Group all points; otherwise no line will show
    geom_point(size=3)
# Dark blue lines, red dots
ggplot(df, aes(x=cond, y=yval)) + 
    geom_line(aes(group=1), colour="#000099") +  # Blue lines
    geom_point(size=3, colour="#CC0000")         # Red dots
```

Instead of changing colors globally, you can map variables to colors – in other words, make the color conditional on a variable, by putting it inside an aes() statement.


```{r}
# Bars: x and fill both depend on cond2
ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity")

```




```{r}
# Bars with other dataset; fill depends on cond2
ggplot(df2, aes(x=cond1, y=yval)) + 
    geom_bar(aes(fill=cond2),   # fill depends on cond2
             stat="identity",
             colour="black",    # Black outline for all
             position=position_dodge()) # Put bars side-by-side instead of stacked


```



```{r}

# Lines and points; colour depends on cond2
ggplot(df2, aes(x=cond1, y=yval)) + 
    geom_line(aes(colour=cond2, group=cond2)) + # colour, group both depend on cond2
    geom_point(aes(colour=cond2),               # colour depends on cond2
               size=3)                          # larger points, different shape
## Equivalent to above; but move "colour=cond2" into the global aes() mapping
# ggplot(df2, aes(x=cond1, y=yval, colour=cond2)) + 
#    geom_line(aes(group=cond2)) +
#    geom_point(size=3)

```
## A colorblind-friendly palette:


```{r}
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# To use for fills, add
  scale_fill_manual(values=cbPalette)

# To use for line and point colors, add
  scale_colour_manual(values=cbPalette)
```

```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) +
    geom_point()
```

```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) +
    geom_point() +
    scale_color_manual(values=c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7"))
```


```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) +
    geom_point() +
    scale_color_manual(values=c("Fair"="#E69F00", "Good"="#56B4E9", "Premium"="#009E73", "Ideal"="#F0E442", "Very Good"="#0072B2"))
```

## Setting luminance and saturation (chromaticity)
```{r}
# Use luminance=45, instead of default 65
ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity") +
    scale_fill_hue(l=40)


```



```{r}
# Reduce saturation (chromaticity) from 100 to 50, and increase luminance
ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity") +
    scale_fill_hue(c=45, l=80)

# Note: use scale_colour_hue() for lines and points

```


```{r}

ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity") +
    scale_fill_brewer()


```




```{r}

ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity") +
    scale_fill_brewer(palette="Set1")


```


```{r}

ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity") +
    scale_fill_brewer(palette="Spectral")

# Note: use scale_colour_brewer() for lines and points
```


### Palettes: manually-defined
```{r}
ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity") + 
    scale_fill_manual(values=c("red", "blue", "green"))


```


```{r}
ggplot(df, aes(x=cond, y=yval, fill=cond)) + geom_bar(stat="identity") + 
    scale_fill_manual(values=c("#CC6666", "#9999CC", "#66CC99"))

# Note: use scale_colour_manual() for lines and points
```

## Continuous colors:

```{r}
# Generate some data
set.seed(133)
df <- data.frame(xval=rnorm(50), yval=rnorm(50))

# Make color depend on yval
ggplot(df, aes(x=xval, y=yval, colour=yval)) + geom_point()


```

```{r}
# Use a different gradient
ggplot(df, aes(x=xval, y=yval, colour=yval)) + geom_point() + 
    scale_colour_gradientn(colours=rainbow(4))
```

[learn more](http://www.cookbook-r.com/Graphs/Colors_%28ggplot2%29/)

